A Tool for Visualizing Patterns of Spreadsheet Function Combinations

### 2016-04-21 doc/paper.tex
Fixes #42

	Many other tools inspect the structure of formulas but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design \cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic formulae \cite{hermans2012detecting}
	\cite{cunha2012towards} \cite{asavametha2012detecting}, often accompanying them
	with detection tools \cite{abreu2014smelling} and investigations
	\cite{jansen2015code}. For example, Hermans, leveraging her work with the
	aforementioned dataflow diagrams, has worked with assessing common
	inter-worksheet smells, such as feature envy and inappropriate intimacy,
	wherein references between worksheets (rather than cells) in a file suggest a
	problem. \cite{hermans2012detectinginter}. Other tools, like Badam's Excel
	refactoring \cite{badame2012refactoring}, are concerned more with understanding
	the formula in the service of changing them. However, these studies define
	standards of poor design and orient their searches around them. Though
	\toolname supports the search for bad design, as we will see, the tool itself
	takes a much more agnostic approach in presentation, offering examples without
	regard for design quality. \par

@@

		\item [3] \textbf{Promote a qualitative understanding of the patterns.}
		\toolname should supplement its observations with concrete instances of
		relevant formulas from the corpus. Ideally, it should even direct users to the
		exact cell in the spreadsheets where the formula was used, contextualizing the
		functions.

@@

	\subsection{Design Decisions} \label{ssec:decisions}
	Early in the process, we decided that the tree
	form would be a suitable fit to represent the parent-child or caller-callee
	relationships inherent in the data, given the composition of formulas as
	functions and their arguments (which could be yet more functions). By adapting
	this structure to accommodate a broad range of possibilities for nested
	functions, the branching factor depends on both the number of arguments in a
	function and the number of possible functions observed as an argument in a
	function. \par

	Guided by the goals in section A, we faced a number of decision points, a
	sample of which we describe below, before we arrived at the design shown above.
	\begin{itemize} \item \textit{Copied formulas}: Excel allows users to spread a
		formula over an area, repeating the same task in each cell with minor
		adjustments. Without checking for this, the analysis may not reveal the
		functions most commonly used together but rather the formulas most often
		applied to large areas. To combat this, we converted formulas from their native
		A1 format to the relative R1C1, in which copied formulas should be identical,
		and reduced the records to unique R1C1 formulas per sheet. \item
		\textit{Importance of depth}: When a function appears within another, should it
		be analyzed only as a nested function or would it also be valid to analyze the
		nested function on its own?

@@

	\subsection{Implementation Details} We can view the final visualization as the
	product of two discrete processes: \begin{itemize} \item \textit{Collection}:
		Given a set of Excel sheets, the tool, written mostly in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form by which it is recorded. When all formulas have
		been analyzed like this, it produces JSON files for each top-level function in
		the set. \item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily for
		its accessibility: given the JSON, the visualization can be rather simply
		embedded into a webpage accessible through a browser. \end{itemize} A
	demonstration of the tool can be found at [LINK HERE].

@@

	\subsection{Positive Responses} The inclusion of formula examples, grounded in
	and linking back to the originating spreadsheets, proved to be an especial draw
	to the tool. After all, none of the participants professed themselves to be
	experts with Excel, their self-reported proficiencies ranging from "fairly
	familiar" (P2) to "not extremely familiar" (P4); even if they were experts, it
	would likely be unreasonable to expect an evenly distributed familiarity with
	the 100+ function trees available. Nevertheless, the list of actual formulas
	accessible through a node's double-click allowed the users a way to find
	examples of formulas with exactly the configuration they wanted. For example,
	as P1 explored the VLOOKUP function, even though they knew the official API
	more verbosely explained its signature, they said they could use the tool to
	collect more examples than the documentation offers. \par

@@

	The discovery of errors and bad design, whether by purpose or accident,
	produced the more jarring events of the study, and as such, most users
	commented on the tool's ability to locate these problems in spreadsheet design.
	This was often supported by the distinctive visual signatures of certain bad
	practices: multiple arguments produce towers of nodes, deep function nesting
	stretches the tree horizontally, and so on. As such, the participants remarked
	on the problems quickly after opening: P3, exploring the example for a
	27-argument SUM (with SUM's initial appearance pictured left), diagnosed the
	situation with either incompetency or just plain weirdness; P2, meanwhile,
	decided that the spreadsheet's author "was having a bad day" when writing an
	AVERAGE with 18 distinct arguments. Aside from these implicit cues, there are
	also nodes which refer to formula errors, prefixed with \#s, which P4 said
	could be useful if marked well. Furthermore, once these errors are found, the
	participants could then open the offending spreadsheets, discovering that, even
	in context, some of these formulas remained incomprehensible, posing problems
	for the spreadsheet transfer scenarios as outlined by Hermans
	\cite{hermans2011supporting}. The tool, then, aggregates these issues in a
	single place, accessible to either the unguided user and the ones who know
	exactly where to find these issues. \par

>>>>

	Many other tools inspect the structure of formulae but often from the
	perspective of finding bad design.  Following Fowler's initial description of
	code smells in object-oriented design \cite{fowler2009refactoring}, many
	researchers have applied similar observations to spreadsheets and have
	cataloged classes of problematic formulae \cite{hermans2012detecting}
	\cite{cunha2012towards} \cite{asavametha2012detecting}, often accompanying them
	with detection tools \cite{abreu2014smelling} and investigations
	\cite{jansen2015code}. For example, Hermans, leveraging her work with the
	aforementioned dataflow diagrams, has worked with assessing common
	inter-worksheet smells, such as feature envy and inappropriate intimacy,
	wherein references between worksheets (rather than cells) in a file suggest a
	problem. \cite{hermans2012detectinginter}. Other tools, like Badam's Excel
	refactoring \cite{badame2012refactoring}, are concerned more with understanding
	the formula in the service of changing them. However, these studies define
	standards of poor design and orient their searches around them. Though
	\toolname supports the search for bad design, as we will see, the tool itself
	takes a much more agnostic approach in presentation, offering examples without
	regard for design quality. \par

@@

		\item [3] \textbf{Promote a qualitative understanding of the patterns.}
		\toolname should supplement its observations with concrete instances of
		relevant formulae from the corpus. Ideally, it should even direct users to the
		exact cell in the spreadsheets where the formula was used, contextualizing the
		functions.

@@

	\subsection{Design Decisions} \label{ssec:decisions}
	Early in the process, we decided that the tree
	form would be a suitable fit to represent the parent-child or caller-callee
	relationships inherent in the data, given the composition of formulae as
	functions and their arguments (which could be yet more functions). By adapting
	this structure to accommodate a broad range of possibilities for nested
	functions, the branching factor depends on both the number of arguments in a
	function and the number of possible functions observed as an argument in a
	function. \par

	Guided by the goals in section A, we faced a number of decision points, a
	sample of which we describe below, before we arrived at the design shown above.
	\begin{itemize} \item \textit{Copied formulae}: Excel allows users to spread a
		formula over an area, repeating the same task in each cell with minor
		adjustments. Without checking for this, the analysis may not reveal the
		functions most commonly used together but rather the formulae most often
		applied to large areas. To combat this, we converted formulae from their native
		A1 format to the relative R1C1, in which copied formulae should be identical,
		and reduced the records to unique R1C1 formulae per sheet. \item
		\textit{Importance of depth}: When a function appears within another, should it
		be analyzed only as a nested function or would it also be valid to analyze the
		nested function on its own?

@@

	\subsection{Implementation Details} We can view the final visualization as the
	product of two discrete processes: \begin{itemize} \item \textit{Collection}:
		Given a set of Excel sheets, the tool, written mostly in Java, uses Apache
		POI\footnote{https://poi.apache.org/} to identify and iterate over every cell
		containing a valid formula. Afterward, it calls POI's formula parser to break
		the formula text into an ordered set of individual tokens, which the tool then
		parses into the tree-like form by which it is recorded. When all formulae have
		been analyzed like this, it produces JSON files for each top-level function in
		the set. \item \textit{Presentation}: The JSON files, meanwhile, feed into the
		presentation code, implemented in Javascript with much help from the
		visualization library D3\footnote{https://d3js.org/}. We chose D3 primarily for
		its accessibility: given the JSON, the visualization can be rather simply
		embedded into a webpage accessible through a browser. \end{itemize} A
	demonstration of the tool can be found at [LINK HERE].

@@

	\subsection{Positive Responses} The inclusion of formula examples, grounded in
	and linking back to the originating spreadsheets, proved to be an especial draw
	to the tool. After all, none of the participants professed themselves to be
	experts with Excel, their self-reported proficiencies ranging from "fairly
	familiar" (P2) to "not extremely familiar" (P4); even if they were experts, it
	would likely be unreasonable to expect an evenly distributed familiarity with
	the 100+ function trees available. Nevertheless, the list of actual formulae
	accessible through a node's double-click allowed the users a way to find
	examples of formulae with exactly the configuration they wanted. For example,
	as P1 explored the VLOOKUP function, even though they knew the official API
	more verbosely explained its signature, they said they could use the tool to
	collect more examples than the documentation offers. \par

@@

	The discovery of errors and bad design, whether by purpose or accident,
	produced the more jarring events of the study, and as such, most users
	commented on the tool's ability to locate these problems in spreadsheet design.
	This was often supported by the distinctive visual signatures of certain bad
	practices: multiple arguments produce towers of nodes, deep function nesting
	stretches the tree horizontally, and so on. As such, the participants remarked
	on the problems quickly after opening: P3, exploring the example for a
	27-argument SUM (with SUM's initial appearance pictured left), diagnosed the
	situation with either incompetency or just plain weirdness; P2, meanwhile,
	decided that the spreadsheet's author "was having a bad day" when writing an
	AVERAGE with 18 distinct arguments. Aside from these implicit cues, there are
	also nodes which refer to formula errors, prefixed with \#s, which P4 said
	could be useful if marked well. Furthermore, once these errors are found, the
	participants could then open the offending spreadsheets, discovering that, even
	in context, some of these formulae remained incomprehensible, posing problems
	for the spreadsheet transfer scenarios as outlined by Hermans
	\cite{hermans2011supporting}. The tool, then, aggregates these issues in a
	single place, accessible to either the unguided user and the ones who know
	exactly where to find these issues. \par
